Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-04-26 paper.tex
Correct spelling.

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.
Since 2005 it has become evident that a good amount of development in open 
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of 
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies 
of small molecules was created as a resource about chemical structure and nomenclature
by biologists \cite{DeMatos:2009p3839}.
Open workflow systems such as Taverna \cite{Hull:2006p60} and KNIME\cite{WebKNIME}
have opened the door for 
an easy and seamless integration of FOS components into executable workflows 
or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010 \cite{Kuhn:2010p4001}.

@@

Some project have also been adopted by other fields, including biology
and crystallography. A good example
here is Jmol, which has, in the past five years, gone through two major revisions
as part of an accelerated development focus. Starting with Jmol 10.2, released in
April of 2006, Jmol 11.0 added over 100 new features and was released in March of
2007. Three years later, Jmol 12.0 followed with over 700 new features.
Based on the outstanding work of Michael (Miguel) Howard and others, this development
path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
visualization package, including full support for crystallography~\cite{Hanson2010},
efficient delivery of surface data using fast solvent-excluded surface
generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
 display of molecular orbitals from standard basis set/coefficient data,
the inclusion of dynamic minimization using the UFF force field, and
a full implementation of Daylight SMILES and SMART, with extensions to
conformational and biomolecular substructure searching in the form of Jmol
BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
Behind the scenes, Jmol has been essentially completely rewritten. A full
mathematical expression evaluator was added, and the scripting language
was folded into a versatile JavaScript-like command language, including a
complete set of program flow commands, several new data types, and
user-defined variables and functions. An exciting non-chemical application
of Jmol has been its incorporation into the open source Sage
Mathematics Project [http://www.sagemath.org/], where Jmol is being used
to deliver three-dimensional depiction of mathematical functions.

@@

More recently, GGA Software Services (a contract programming company)
released the Indigo toolkit \cite{WebIndigo} and associated software
in 2009 under the GPL. Indigo is written from scratch in C++ and has
high-level wrappers for using it in C, Java, Python, or .NET
environment. Like other toolkits, Indigo provides support for
tetrahedral and cis-trans stereochemistry, 2D coordinate generation,
exact/substructure/SMARTS matching, fingerprint generation (all these
both for molecules and reactions), and canonical SMILES computation.
It also provides some less common functionality, like matching
tautomers and resonance substructures, enumeration of subgraps,
finding maximum common substructure of $N$ input structures, and
enumerating reaction products. Indigo also has its rendering plugin,
which is used in a lot of commercial and open-source projects as the
main 2D chemical rendering engine.

While desktop software has composed the majority of scientific tools since the computer was introduced, the internet continues to change how applications and content are distributed and presented. As such, there is a growing need to present chemical and scientific information in a web-based medium. Recently, a new version of the HTML specification, HTML5 \cite{html5}, defines a well-developed framework for creating native web applications in Javascript. This native framework provides a significant benefit to scientists as the web is an open and free medium to distribute scientific knowledge, ideas and education. It also poses significant new problems, as the cost associated with learning and developing tools for new technologies is not trivial. As such, it is an important goal for the Blue Obelisk to provide such tools and encourage the community to utilize these new technologies to improve products and education. A specific example of this initiative is the ChemDoodle Web Components library \cite{ChemDoodleWeb}, produced by iChemLabs. The ChemDoodle Web Components library is open source under the GPL v3 license with a liberal HTML exception. ChemDoodle Web Components allow the scientist to present publication quality 2D and 3D graphics and animations for chemical structures, reactions and spectra. Beyond graphics, this tool provides a framework for user interaction to create dynamic applications through web browsers, desktop platforms and mobile devices such as the iPhone, iPad and Android devices. It is now easier than ever for scientists to create web-based scientific content, without having to master complicated technologies such as Java applets. iChemLabs is dedicated to funding, developing and supporting the library to ensure that the next generation of scientific applications is easily achievable by academia, government and industry. This goal helps to make sure that the cost of education decreases while using the web to further spread science.

@@

CDK-Taverna \cite{Kuhn:2010p4001} as an open-source cheminformatics workflow solution.
Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharma- cophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc. 
CDK-Taverna has matured to become a freely available and increasingly powerful tool for the biosciences. It was recently ported to 
Taverna version 2.x.
The combination of CDK-Taverna workers with the multitude of workers already published on myexperiment.org by the large and active user community of Taverna enables scientists to quickly build workflows to process a diverse set of data as typically found in today's systems biology scenarios.

@@

Open Source provides a unique opportunity for commercial organizations to work with the community. As opposed to traditional business models which focus on maximizing revenue through monetization of all aspects of software, working with the community allows a business to maximize profits through greater community adoption and popularity. This is an ideal solution for some commercial organizations because only they can afford the significant investment in funding and time that is required to develop a successful open source project. Regardless, there are always challenges to growing a project, and there are specific issues that complicate making a business out of open source software in the scientific industry. A goal of the Blue Obelisk is to reduce these challenges and bring commercial organizations and the community together to benefit science.

@@

Successful open projects require the dedication of a number of individuals to ensure that the product is continually developed, supported and funded. In many non-commercial cases, an author will produce an open project as a hobby, and stop work on the project before the development reaches a certain quality or community interest reaches a critical mass. This is because the hobbyist eventually needs to focus on income or he/she has a change of priorities. On the other hand, commercial organizations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption. In return, that commercial organization gains recognition from that product and is the first choice for commercial support, consulting and custom development. This is the primary goal of commercial open source, as selling licenses for an open product is very unlikely to be feasible.

In the scientific industry, it is incredibly difficult to start successful open projects. This is because the expertise needed to contribute to such a project requires a higher degree of education and the population of end users is much smaller than other industries, such as entertainment. Because of this, there is an increased barrier to starting open projects in the scientific industry. The Blue Obelisk attempts to solve this issue, by uniting those interested in open source and by helping to increase awareness for new open projects. Commercial organizations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.

For instance, iChemLabs distributes several scientific software solutions under various open and proprietary licensing. The license is chosen to best ensure the success of the project. iChemLabs released the ChemDoodle Web Components library under the GPL v3 license with a liberal HTML exception. iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilize the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content. Other open source developers can benefit from this library freely and expect continued development, while commercial organizations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market. Of course, no one will need to waste time rewriting functionality that is already provided with quality. Several open projects have already incorporated the ChemDoodle Web Components, such as iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.

@@

Performance. There is a famous quote by Knuth that "Premature
optimization is root of all evil (or something)". Given that compute
time is cheap, ....

@@

Recently proposed CurlySMILES\cite{CurlySMILES} is an extension of the
SMILES notation, which allows to define crystals structures, polymers,
electron delocalization charges, molecule interations, and many other
features absent in the initial SMILES specification. There has been
discussions about including parts of the CurlySMILES notation into
OpenSMILES, especially polymers.

@@

\subsubsection*{QSAR-ML}
The field of QSAR has long been hampered by the lack of open standards, which makes it difficult to share and reproduce descriptor calculations and analyses. QSAR-ML was recently proposed as an open standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}. A dataset in QSAR-ML includes the chemical structures (preferable described in CML) with InChI to protect integrity, chemical descriptors by linking to the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values, units, and versioned descriptor implementations to allow for integrating several descriptor software in the same calculation. Hence, a dataset described in QSAR-ML is completely reproducible. To allow for easy setup of QSAR-ML compliant datasets, a plugin for Bioclipse was created with graphical interfaces that can be used to set up QSAR datasets and perform calculations. Descriptor implementations were initially available from CDK and JoeLIB, as well as via remote web services such as XMPP~\cite{Wagener:2009uq}.

@@

A core requirement for chemical structure databases and chemical
registration systems in general is the notion of structure
standardization.  That is,  for a given input structure, multiple
representations should be converted to one canonical form. 
Structure canonicalization routines partially address this aspect,
converting multiple alternative topologies to a single canonical
form. However, the problem of standardization is broader than just
topological canonicalization. Features that must be considered include
\begin{itemize}
\item topological canonicalization
\item handling of charges
\item tautomer enumeration and canonicalization
\item normalization of functional groups
\end{itemize}
Currently, most of the individual components of a `standardization
pipeline' can be implemented using BO tools. The larger problem is
that there is no agreed upon list of steps for a standardization
process. While some specifications have been published (e.g., Pubchem)
and some standardization services and tools are available (Pubchem
provides an online service to standardise molecules and the NCGC
provides a stand alone tool) each group has their own set of rules. A
common reference specification for standardization would be of immense
value in interoperability between structure repositories as well as
between toolkits (though the latter is still confounded by differences
in lower level cheminformatic features such as aromaticity models).

@@

Importantly, publishing data as CC0 is become easier now that
website are becoming available to simplify publishing data. Two
projects that can be mentioned in this context are FigShare
(http://figshare.com/), where the data behind unpublished figures
can be hosted, and Dryad (http://datadryad.org/) where data
behind publications can be hosted. Initiatives like this make
it possible to host small amounts of data, and those combinded
are expected to become soon a substantial knowledge base.

@@

\section*{Other areas of activity}
Web services? XMPP services? (NIH, CDK and Chemspider)

@@

One recent development
that attempts to address this issue was the establishment in Apr 2010 of a
question-and-answer (QA)
website related to Blue Obelisk projects and themes at
http://blueobelisk.shapado.com (see Figure~3). This is a website in the
style of Stack Overflow that encourages high quality answers (and
questions) through the use of a voting system. In the year since it
was established, over
200 users have registered, many of whom had no previous involvement
with the Blue Obelisk, showing that the QA website is a nice complement
to earlier existing channels of communication.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.
Since 2005 it has become evident that a good amount of development in open 
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of 
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies 
of small molecules was created as a resource about chemical structure and nomenclature
by biologists \cite{DeMatos:2009p3839}.
Open workflow systems such as Taverna \cite{Hull:2006p60} and KNIME\cite{WebKNIME}
have opened the door for 
an easy and seamless integration of BO components into executable workflows 
or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010 \cite{Kuhn:2010p4001}.

@@

Some project have also been adopted by other fields, including biology
and crystallography. A good example
here is Jmol, which has, in the past five years, gone through two major revisions
as part of an accelerated development focus. Starting with Jmol 10.2, released in
April of 2006, Jmol 11.0 added over 100 new features and was released in March of
2007. Three years later, Jmol 12.0 followed with over 700 new features.
Based on the outstanding work of Michael (Miguel) Howard and others, this development
path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
visualisation package, including full support for crystallography~\cite{Hanson2010},
efficient delivery of surface data using fast solvent-excluded surface
generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
 display of molecular orbitals from standard basis set/coefficient data,
the inclusion of dynamic minimisation using the UFF force field, and
a full implementation of Daylight SMILES and SMART, with extensions to
conformational and biomolecular substructure searching in the form of Jmol
BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
Behind the scenes, Jmol has been essentially completely rewritten. A full
mathematical expression evaluator was added, and the scripting language
was folded into a versatile JavaScript-like command language, including a
complete set of program flow commands, several new data types, and
user-defined variables and functions. An exciting non-chemical application
of Jmol has been its incorporation into the open source Sage
Mathematics Project [http://www.sagemath.org/], where Jmol is being used
to deliver three-dimensional depiction of mathematical functions.

@@

More recently, GGA Software Services (a contract programming company)
released the Indigo toolkit \cite{WebIndigo} and associated software
in 2009 under the GPL. Indigo is written from scratch in C++ and has
high-level wrappers for using it in C, Java, Python, or .NET
environment. Like other toolkits, Indigo provides support for
tetrahedral and cis-trans stereochemistry, 2D coordinate generation,
exact/substructure/SMARTS matching, fingerprint generation (all these
both for molecules and reactions), and canonical SMILES computation.
It also provides some less common functionality, like matching
tautomers and resonance substructures, enumeration of subgraphs,
finding maximum common substructure of $N$ input structures, and
enumerating reaction products. Indigo also has its rendering plugin,
which is used in a lot of commercial and open-source projects as the
main 2D chemical rendering engine.

While desktop software has composed the majority of scientific tools since the computer was introduced, the internet continues to change how applications and content are distributed and presented. As such, there is a growing need to present chemical and scientific information in a web-based medium. Recently, a new version of the HTML specification, HTML5 \cite{html5}, defines a well-developed framework for creating native web applications in JavaScript. This native framework provides a significant benefit to scientists as the web is an open and free medium to distribute scientific knowledge, ideas and education. It also poses significant new problems, as the cost associated with learning and developing tools for new technologies is not trivial. As such, it is an important goal for the Blue Obelisk to provide such tools and encourage the community to utilise these new technologies to improve products and education. A specific example of this initiative is the ChemDoodle Web Components library \cite{ChemDoodleWeb}, produced by iChemLabs. The ChemDoodle Web Components library is open source under the GPL v3 license with a liberal HTML exception. ChemDoodle Web Components allow the scientist to present publication quality 2D and 3D graphics and animations for chemical structures, reactions and spectra. Beyond graphics, this tool provides a framework for user interaction to create dynamic applications through web browsers, desktop platforms and mobile devices such as the iPhone, iPad and Android devices. It is now easier than ever for scientists to create web-based scientific content, without having to master complicated technologies such as Java applets. iChemLabs is dedicated to funding, developing and supporting the library to ensure that the next generation of scientific applications is easily achievable by academia, government and industry. This goal helps to make sure that the cost of education decreases while using the web to further spread science.

@@

CDK-Taverna \cite{Kuhn:2010p4001} as an open-source cheminformatics workflow solution.
Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharmacophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc. 
CDK-Taverna has matured to become a freely available and increasingly powerful tool for the biosciences. It was recently ported to 
Taverna version 2.x.
The combination of CDK-Taverna workers with the multitude of workers already published on myexperiment.org by the large and active user community of Taverna enables scientists to quickly build workflows to process a diverse set of data as typically found in today's systems biology scenarios.

@@

Open Source provides a unique opportunity for commercial organisations to work with the community. As opposed to traditional business models which focus on maximising revenue through monetisation of all aspects of software, working with the community allows a business to maximise profits through greater community adoption and popularity. This is an ideal solution for some commercial organisations because only they can afford the significant investment in funding and time that is required to develop a successful open source project. Regardless, there are always challenges to growing a project, and there are specific issues that complicate making a business out of open source software in the scientific industry. A goal of the Blue Obelisk is to reduce these challenges and bring commercial organisations and the community together to benefit science.

@@

Successful open projects require the dedication of a number of individuals to ensure that the product is continually developed, supported and funded. In many non-commercial cases, an author will produce an open project as a hobby, and stop work on the project before the development reaches a certain quality or community interest reaches a critical mass. This is because the hobbyist eventually needs to focus on income or he/she has a change of priorities. On the other hand, commercial organisations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption. In return, that commercial organisation gains recognition from that product and is the first choice for commercial support, consulting and custom development. This is the primary goal of commercial open source, as selling licenses for an open product is very unlikely to be feasible.

In the scientific industry, it is incredibly difficult to start successful open projects. This is because the expertise needed to contribute to such a project requires a higher degree of education and the population of end users is much smaller than other industries, such as entertainment. Because of this, there is an increased barrier to starting open projects in the scientific industry. The Blue Obelisk attempts to solve this issue, by uniting those interested in open source and by helping to increase awareness for new open projects. Commercial organisations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.

For instance, iChemLabs distributes several scientific software solutions under various open and proprietary licensing. The license is chosen to best ensure the success of the project. iChemLabs released the ChemDoodle Web Components library under the GPL v3 license with a liberal HTML exception. iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilise the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content. Other open source developers can benefit from this library freely and expect continued development, while commercial organisations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market. Of course, no one will need to waste time rewriting functionality that is already provided with quality. Several open projects have already incorporated the ChemDoodle Web Components, such as iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.

@@

Performance. There is a famous quote by Knuth that "Premature
optimisation is root of all evil (or something)". Given that compute
time is cheap, ....

@@

Recently proposed CurlySMILES\cite{CurlySMILES} is an extension of the
SMILES notation, which allows to define crystals structures, polymers,
electron delocalisation charges, molecule interactions, and many other
features absent in the initial SMILES specification. There has been
discussions about including parts of the CurlySMILES notation into
OpenSMILES, especially polymers.

@@

\subsubsection*{QSAR-ML}
The field of QSAR has long been hampered by the lack of open standards, which makes it difficult to share and reproduce descriptor calculations and analyses. QSAR-ML was recently proposed as an open standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}. A dataset in QSAR-ML includes the chemical structures (preferable described in CML) with InChI to protect integrity, chemical descriptors by linking to the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values, units, and versioned descriptor implementations to allow for integrating several descriptor software in the same calculation. Hence, a dataset described in QSAR-ML is completely reproducible. To allow for easy setup of QSAR-ML compliant datasets, a plugin for Bioclipse was created with graphical interfaces that can be used to set up QSAR datasets and perform calculations. Descriptor implementations were initially available from CDK and JOELib, as well as via remote web services such as XMPP~\cite{Wagener:2009uq}.

@@

A core requirement for chemical structure databases and chemical
registration systems in general is the notion of structure
standardisation.  That is,  for a given input structure, multiple
representations should be converted to one canonical form. 
Structure canonicalisation routines partially address this aspect,
converting multiple alternative topologies to a single canonical
form. However, the problem of standardisation is broader than just
topological canonicalisation. Features that must be considered include
\begin{itemize}
\item topological canonicalisation
\item handling of charges
\item tautomer enumeration and canonicalisation
\item normalisation of functional groups
\end{itemize}
Currently, most of the individual components of a `standardisation
pipeline' can be implemented using BO tools. The larger problem is
that there is no agreed upon list of steps for a standardisation
process. While some specifications have been published (e.g., PubChem)
and some standardisation services and tools are available (PubChem
provides an online service to standardise molecules and the NCGC
provides a stand alone tool) each group has their own set of rules. A
common reference specification for standardisation would be of immense
value in interoperability between structure repositories as well as
between toolkits (though the latter is still confounded by differences
in lower level cheminformatic features such as aromaticity models).

@@

Importantly, publishing data as CC0 is become easier now that
website are becoming available to simplify publishing data. Two
projects that can be mentioned in this context are FigShare
(http://figshare.com/), where the data behind unpublished figures
can be hosted, and Dryad (http://datadryad.org/) where data
behind publications can be hosted. Initiatives like this make
it possible to host small amounts of data, and those combined
are expected to become soon a substantial knowledge base.

@@

\section*{Other areas of activity}
Web services? XMPP services? (NIH, CDK and ChemSpider)

@@

One recent development
that attempts to address this issue was the establishment in Apr 2010 of a
question-and-answer (Q\&A)
website related to Blue Obelisk projects and themes at
http://blueobelisk.shapado.com (see Figure~3). This is a website in the
style of Stack Overflow that encourages high quality answers (and
questions) through the use of a voting system. In the year since it
was established, over
200 users have registered, many of whom had no previous involvement
with the Blue Obelisk, showing that the Q\&A website is a nice complement
to earlier existing channels of communication.
